---
output:
  html_document:
    theme: spacelab
---

```{r}
library(SummarizedExperiment)
library(jaffelab)
library(sessioninfo)
library(dplyr)
library(reshape2)
```


## Swap samples identified in Brain Swap project
5: Br2260: 9373406026_R01C01
$`Lims Phenotype Info`
     BrNum sex agedeath primarydx race
854 Br2260   F    45.51   Bipolar CAUC
972 Br2473   F    52.61   Bipolar CAUC

$BrNum
[1] "Br2260" "Br2473"

$`Flag index`
[1] 5 9

$`Genotype data`
                    ID  BrNum bestCor bestCorBr           bestCorID nBr_samp genoSex bestcorSex
189  9373408026_R01C01 Br2473   0.271    Br8067 201962800059_R05C01        1       F          M
1241 9373406026_R01C01 Br2260   0.251    Br1546   4572348603_R01C02        1       F          F

$`RNA data`
    cor genoBr        genoSample  rnaBr        rnaSample genoChip rnaRegion     rnaDataset genoSex rnaSex
1 0.920 Br2260 9373406026_R01C01 Br2473 R14077_HCTYLBBXX       5M  Amygdala psychENCODE_BP       F      F
2 0.921 Br2260 9373406026_R01C01 Br2473 R14296_H7JLCBBXX       5M      sACC psychENCODE_BP       F      F
3 0.899 Br2473 9373408026_R01C01 Br2260 R14071_HF3JYBBXX       5M  Amygdala psychENCODE_BP       F      F
4 0.951 Br2473 9373408026_R01C01 Br2260 R14290_H7L3FBBXX       5M      sACC psychENCODE_BP       F      F

10: Br2486: 9373408026_R02C01
$`Lims Phenotype Info`
     BrNum sex agedeath primarydx race
864 Br2275   F    45.03   Bipolar CAUC
979 Br2486   M    46.73    Schizo   AA

$BrNum
[1] "Br2275" "Br2486"

$`Flag index`
[1]  10 100 101

$`Genotype data`
                      ID  BrNum bestCor bestCorBr           bestCorID nBr_samp genoSex bestcorSex
48     9373408026_R02C01 Br2486   0.995    Br2275 201398400130_R06C01        1       F          F
1892   9373406026_R02C01 Br2275   0.324    Br5085   8501651024_R04C01        2       M          F
1958 201398400130_R06C01 Br2275   0.995    Br2486   9373408026_R02C01        2       F          F

$`RNA data`
    cor genoBr          genoSample  rnaBr        rnaSample   genoChip rnaRegion     rnaDataset genoSex rnaSex
1 0.961 Br2275 201398400130_R06C01 Br2275 R13997_H7JHNBBXX 2-5-8-v1-3  Amygdala psychENCODE_BP       F      F
2 0.956 Br2275 201398400130_R06C01 Br2275 R14222_H7JHNBBXX 2-5-8-v1-3      sACC psychENCODE_BP       F      F
3 0.956 Br2486   9373408026_R02C01 Br2275 R14222_H7JHNBBXX         5M      sACC psychENCODE_BP       F      F
4 0.961 Br2486   9373408026_R02C01 Br2275 R13997_H7JHNBBXX         5M  Amygdala psychENCODE_BP       F      F

## List swap samples
```{r}
swap_ID <- c("R14077_HCTYLBBXX","R14296_H7JLCBBXX","R14071_HF3JYBBXX","R14290_H7L3FBBXX","R13997_H7JHNBBXX","R14222_H7JHNBBXX")
swap_case <- c(rep("swap_simple",4),rep("swap_complex", 2))
swap_df <- data.frame(swap_ID, swap_case)
colnames(swap_df) <- c("SAMPLE_ID", "SwapCase")
swap_df
```

## Load Data
fastq files found in /dcl02/lieber/ajaffe/Nina/Zandi/data
```{r}
## zandi/hyde bipolar
load("/dcl01/lieber/ajaffe/lab/zandiHyde_bipolar_rnaseq/data/zandiHypde_bipolar_rseGene_n511.rda", verbose = TRUE)
pd <- colData(rse_gene)[,c("SAMPLE_ID","RNum","BrNum","BrainRegion","Sex","AgeDeath","Race","PrimaryDx","RIN")]
head(pd)
```

## Select samples
### Get swap samples PD
All 3 individuals the samples are from are F, 45+, Cauc, and Bipolar
```{r}
pd_swap <- merge(swap_df, pd)
(pd_swap <- as_tibble(pd_swap))
```

### Select other (no-swap) samples 
want 40 samples in the final data set, after 2 samples from the complex swap are dropped
balance across Dx, Region, Sex, similar mean in Age
```{r}
#all other samples
pd_other <- subset(pd, !SAMPLE_ID %in% swap_ID)
dim(pd_other)
#check for all adult samples
summary(pd_other$AgeDeath)
#  Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
# 17.37   35.34   47.09   47.18   56.05   86.76 
subset(pd_other, PrimaryDx == "Bipolar" & Sex == "F" & BrainRegion == "Amygdala")

# set seed for reproducibility
set.seed(456)

#randomy select 5 cases from each group
pd_other40 <- pd_other %>% as_tibble() %>% 
    filter(PrimaryDx != "Other") %>%
    mutate(SwapCase = "no_swap") %>%
    group_by(BrainRegion,Sex, PrimaryDx) %>%
    sample_n(5) %>%
    ungroup()

pd_other40 %>% group_by(PrimaryDx) %>% 
    summarise(mean(AgeDeath),median(AgeDeath))
```

### replace four no_swap samples with the simple_swap cases, merge all tables together
```{r}
pd_swap %>% filter(SwapCase == "swap_simple") %>% count(Sex, PrimaryDx, BrainRegion)
#   Sex   PrimaryDx BrainRegion     n
#   <fct> <fct>     <fct>       <int>
# 1 F     Bipolar   Amygdala        2
# 2 F     Bipolar   sACC            2

# need to Drop 2 F, Bipolar samples from each region
pd_drop <- pd_other40 %>% 
    filter(PrimaryDx == "Bipolar", Sex == "F") %>%
    group_by(BrainRegion) %>%
    sample_n(2)

#merge tables    
(pd_example <- pd_other40 %>%
    anti_join(pd_drop) %>%
    rbind(pd_swap))

pd_example %>% filter(SwapCase != "swap_complex") %>%
    group_by(PrimaryDx) %>%
    summarise(mean(AgeDeath),median(AgeDeath))
```

## Match Up With file manifest
```{r}
manifest <- read.delim("/dcl01/lieber/ajaffe/lab/zandiHyde_bipolar_rnaseq/preprocessed_data/.samples_unmerged.manifest", header = FALSE)
head(manifest)
#change to dcl02, same relative path after
manifest<- data.frame(lapply(manifest, function(x) {gsub("dcl01", "dcl02", x)}))
head(manifest)

manifest2 <- manifest[,c("V1","V3","V5")] %>%
    melt(id.vars = "V5") %>% 
    rename(SAMPLE_ID = V5) %>%
    group_by(SAMPLE_ID) %>%
    summarise(FASTQpath = list(value))

pd_example <- pd_example %>% left_join(manifest2, by = "SAMPLE_ID") %>% as.data.frame()
head(pd_example)
save(pd_example, file = "pd_example.Rdata")
```

## Write a subsetted manifest in the format accepted by SPEAQeasy
```{r}
#  These will be columns in the subsetted manifest
R1 = list()
R2 = list()
ids = list()

#  Loop through each id in the pd_example table
counter = 1
for (i in 1:length(pd_example$FASTQpath)) {
    num_files = length(pd_example$FASTQpath[[i]])
    stopifnot((num_files %% 2) == 0) # data is paired
    
    #  Ultimately produce one row for each pair of files of a given id
    for (j in 1:(num_files / 2)) {
        R1[[counter]] = pd_example$FASTQpath[[i]][j]
        R2[[counter]] = pd_example$FASTQpath[[i]][j + num_files / 2]
        ids[[counter]] = pd_example$SAMPLE_ID[[i]]
        
        counter = counter + 1
    }
}

#  Format as a vector of strings, which we will write to the 'sample_selection'
#  folder in the repository
out_manifest = paste(unlist(R1), 0, unlist(R2), 0, unlist(ids), sep='\t')
writeLines(out_manifest, 'samples.manifest')
```
