---
title: 'SPEAQeasy Differential Expression Analysis'
date: 'July 10, 2020'
output:
  BiocStyle::html_document:
    toc: true
  pdf_document:
    toc: true
    number_sections: true
---

```{r 'setup', echo = FALSE, warning = FALSE, message = FALSE}
timestart <- Sys.time()

## Bib setup
library('knitcitations')

## Load knitcitations with a clean bibliography
cleanbib()
cite_options(hyperlink = 'to.doc', citation_format = 'text', style = 'html')

## Write bibliography information
bibs <- c(
    SummarizedExperiment = citation('SummarizedExperiment'),
    recount = citation('recount'),
    edgeR = citation('edgeR'),
    limma = citation('limma'),
    jaffelab = citation('jaffelab'),
    RColorBrewer = citation('RColorBrewer'),
    clusterProfiler = citation('clusterProfiler'),
    org.Hs.eg.db = citation('org.Hs.eg.db'),
    pheatmap = citation('pheatmap'),R = citation(),
    voom = RefManageR::BibEntry('article', key = 'voom', author = 'CW Law and Y Chen and W Shi and GK Smyth', year = '2014', title = 'Voom: precision weights unlock linear model analysis tools for RNA-seq read counts', journal = 'Genome Biology', volume = '15', pages = 'R29'),
    here = citation("here")
)

write.bibtex(bibs, file = 'de_analysis.bib')
bib <- read.bibtex('de_analysis.bib')

## Assign short names
names(bib) <- names(bibs)
```

# Analysis

The following analysis explores a `RangedSummarizedExperiment` object from the SPEAQeasy pipeline. Note that we will use a modified version of the object, which resolved sample identity issues which were present in the raw output from SPEAQeasy. This object also includes phenotype data added after resolving identity issues. Though SPEAQeasy produces objects for several feature types (genes, exons, exon-exon junctions), we will demonstrate an example analysis for just genes. We will perform differential expression across some typical variables of interest (e.g. sex, age, race) and show how to perform principal component analysis (PCA) and visualize findings with plots.

## Load required libraries

```{r loadpackages, message = FALSE, warning = FALSE}
library("SummarizedExperiment")
library("recount")
library("edgeR")
library("limma")
library("jaffelab") # GitHub: LieberInstitute/jaffelab
library("RColorBrewer")
library("clusterProfiler")
library("org.Hs.eg.db")
library("pheatmap")
library("here")
```

## Load data and prepare directories to place outputs in

For those who ran SPEAQeasy from the example FASTQ data set, the `RangedSummarizedExperiment` will have a different path, as specified with the `--output` flag. Set the working directory according to where the `SPEAQeasy-example` repository is cloned.

```{r loaddata}
# Load the RSE gene object
load(here("rse_speaqeasy.RData"), verbose = TRUE)

# Create directories to organize outputs from this analysis
dir.create("pdfs", showWarnings = FALSE)
dir.create("tables", showWarnings = FALSE)
dir.create("rdas", showWarnings = FALSE)
```



## cell PCs
col_names = c('trimmed', 'numReads', 'numMapped', 'numUnmapped', 
              'overallMapRate', 'concordMapRate', 'totalMapped', 'mitoMapped',
              'mitoRate', 'totalAssignedGene')
cellPca = prcomp(as.data.frame(colData(rse_gene)[,col_names]))
rse_gene$PC = cellPca$x[,1]
getPcaVars(cellPca)[1] # 87.3
round(cellPca$rot[,1],3)
# fetal quiescent and adult neuron increase




## filter for expressed
rse_gene = rse_gene[rowMeans(getRPKM(rse_gene,"Length")) > 0.2,]

##############
## metrics ###


## check if ratios of cell changed by batch
pdf(file = "DE_analysis/pdfs/Region_Race_cellcheck.pdf")
boxplot(rse_gene$rRNA_rate ~ rse_gene$BrainRegion,xlab="")
boxplot(rse_gene$mitoRate ~ rse_gene$BrainRegion,xlab="")
boxplot(rse_gene$gene_Assigned ~ rse_gene$BrainRegion,xlab="")
boxplot(rse_gene$mitoRate ~ rse_gene$Race,las=3,xlab="")
boxplot(rse_gene$gene_Assigned ~ rse_gene$Race,las=3,xlab="")
dev.off()

#### explore human
geneExprs = log2(getRPKM(rse_gene,"Length")+1)
pca = prcomp(t(geneExprs))
pca_vars = getPcaVars(pca)
pca_vars_lab = paste0("PC", seq(along=pca_vars), ": ",
	pca_vars, "% Var Expl")


##########
pdf("DE_analysis/pdfs/PCA_plotsExprs.pdf",w=9)
par(mar=c(8,6,2,2),cex.axis=1.8,cex.lab=1.8)
palette(brewer.pal(4,"Dark2"))

## pc1 vs pc2
plot(pca$x, pch=21, bg=factor(rse_gene$PrimaryDx),cex=1.2,
	xlab = pca_vars_lab[1], ylab = pca_vars_lab[2])
legend("bottomleft", levels(rse_gene$PrimaryDx), col=1:2, pch=15,cex=2)


## by line
for(i in 1:10) {
	boxplot(pca$x[,i] ~ rse_gene$Sex,
		ylab=pca_vars_lab[i], las = 3,xlab="Sex",outline=FALSE)
	points( pca$x[,i] ~ jitter(as.numeric(factor(rse_gene$Sex))),
		pch = 21, bg = rse_gene$PrimaryDx,cex=1.2)
}

## by experiment
for(i in 1:10) {
	boxplot(pca$x[,i] ~ rse_gene$Race,
		ylab=pca_vars_lab[i], las = 3,xlab="Race",outline=FALSE)
	points( pca$x[,i] ~ jitter(as.numeric(factor(rse_gene$Race))),
		pch = 21, bg = rse_gene$PrimaryDx,cex=1.2)
}
dev.off()


####################
## modeling ########
####################

dge = DGEList(counts = assays(rse_gene)$counts,
	genes = rowData(rse_gene))
dge = calcNormFactors(dge)

## mean-variance
mod = model.matrix(~PrimaryDx + PC + BrainRegion,
	data=colData(rse_gene))
pdf(file = "DE_analysis/pdfs/vGene.pdf")
vGene = voom(dge,mod,plot=TRUE)
dev.off()

gene_dupCorr = duplicateCorrelation(vGene$E, mod,
	block=colData(rse_gene)$SAMPLE_ID)
save(gene_dupCorr, file = "DE_analysis/rdas/gene_dupCorr_neurons.rda")

fitGeneDupl = lmFit(vGene,
	correlation=gene_dupCorr$consensus.correlation,
	block=colData(rse_gene)$SAMPLEID)

ebGeneDupl = eBayes(fitGeneDupl)
outGeneDupl = topTable(ebGeneDupl,coef=2,
	p.value = 1,number=nrow(rse_gene),sort="none")

pdf(file = "DE_analysis/pdfs/hist_pval.pdf")
hist(outGeneDupl$P.Value)
dev.off()
table(outGeneDupl$adj.P.Val < 0.05)
table(outGeneDupl$adj.P.Val < 0.1)

sigGeneDupl =  topTable(ebGeneDupl,coef=2,
	p.value = 0.1,number=nrow(rse_gene))

sigGeneDupl[,c("Symbol","logFC", "P.Value","AveExpr")]
sigGeneDupl[sigGeneDupl$logFC > 0,c("Symbol","logFC", "P.Value")]
sigGeneDupl[sigGeneDupl$logFC <  0,c("Symbol","logFC", "P.Value")]

write.csv(outGeneDupl, file = "DE_analysis/tables/de_stats_allExprs.csv")
write.csv(sigGeneDupl, file = "DE_analysis/tables/de_stats_fdr10_sorted.csv")

###################
## check plots ####
###################

exprs = vGene$E[rownames(sigGeneDupl),]
#exprsClean = cleaningY(exprs, mod, 2)


### make boxplots
# cleanGeneExprs = cleaningY(geneExprs, mod[,!is.na(eBGene$p.value[1,])], P=3)


pdf("DE_analysis/pdfs/DE_boxplots_byDiagnosis.pdf",w=10)
par(mar=c(8,6,4,2),cex.axis=1.8,cex.lab=1.8, cex.main=1.8)
palette(brewer.pal(4,"Dark2"))
for(i in 1:nrow(sigGeneDupl)) {
	yy = exprs[i,]
	boxplot(yy ~ rse_gene$PrimaryDx, outline=FALSE,
		ylim=range(yy), ylab="Normalized log2 Exprs", xlab="",
		main = paste(sigGeneDupl$Symbol[i], "-", sigGeneDupl$gencodeID[i]))
	points(yy ~ jitter(as.numeric(rse_gene$PrimaryDx)),
		pch = 21, bg= rse_gene$PrimaryDx,cex=1.3)
	ll = ifelse(sigGeneDupl$logFC[i] > 0, "topleft", "topright")
	legend(ll, paste0("p=", signif(sigGeneDupl$P.Value[i],3)), cex=1.3)
}
dev.off()

#### RPKM #####
e = geneExprs[rownames(sigGeneDupl),]

pdf("DE_analysis/pdfs/DE_boxplots_byGenome_log2RPKM.pdf",w=10)
par(mar=c(8,6,4,2),cex.axis=1.8,cex.lab=1.8, cex.main=1.8)
palette(brewer.pal(4,"Dark2"))
for(i in 1:nrow(sigGeneDupl)) {
	yy = e[i,]
	boxplot(yy ~ rse_gene$PrimaryDx, las=3,outline=FALSE,
		ylim=range(yy), ylab="log2(RPKM+1)", xlab="",
		main = paste(sigGeneDupl$Symbol[i], "-", sigGeneDupl$gencodeID[i]))
	points(yy ~ jitter(as.numeric(rse_gene$PrimaryDx)),
		pch = 21, bg= rse_gene$PrimaryDx,cex=1.3)
	ll = ifelse(sigGeneDupl$logFC[i] > 0, "topleft", "topright")
	legend(ll, paste0("p=", signif(sigGeneDupl$P.Value[i],3)), cex=1.3)
}
dev.off()


#################### no rat astrocyte differences

### gene ontology


## get significant genes by sign
sigGene = outGeneDupl[outGeneDupl$P.Value < 0.005,]
sigGeneList = split(as.character(sigGene$EntrezID), sign(sigGene$logFC))
sigGeneList = lapply(sigGeneList, function(x) x[!is.na(x)])
geneUniverse = as.character(outGeneDupl$EntrezID)
geneUniverse = geneUniverse[!is.na(geneUniverse)]

## do GO and KEGG
goBP_Adj <- compareCluster(sigGeneList, fun = "enrichGO",
	universe = geneUniverse, OrgDb = org.Hs.eg.db,
	ont = "BP", pAdjustMethod = "BH", pvalueCutoff  = 1,
	qvalueCutoff  = 1,	readable= TRUE)

goMF_Adj <- compareCluster(sigGeneList, fun = "enrichGO",
	universe = geneUniverse, OrgDb = org.Hs.eg.db,
	ont = "MF", pAdjustMethod = "BH", pvalueCutoff  = 1,
	qvalueCutoff  = 1,	readable= TRUE)

goCC_Adj <- compareCluster(sigGeneList, fun = "enrichGO",
	universe = geneUniverse, OrgDb = org.Hs.eg.db,
	ont = "CC", pAdjustMethod = "BH", pvalueCutoff  = 1,
	qvalueCutoff  = 1,	readable= TRUE)

kegg_Adj <- compareCluster(sigGeneList, fun = "enrichKEGG",
	universe = geneUniverse,  pAdjustMethod = "BH",
	pvalueCutoff  = 1, qvalueCutoff  = 1)

save(goBP_Adj, goCC_Adj, goMF_Adj, kegg_Adj,
	file = "DE_analysis/rdas/gene_set_objects_p005.rda")

goList = list(BP = goBP_Adj, MF = goMF_Adj, CC = goCC_Adj, KEGG = kegg_Adj)
goDf = dplyr::bind_rows(lapply(goList, as.data.frame), .id = "Ontology")
goDf = goDf[order(goDf$pvalue),]

write.csv(goDf, file = "DE_analysis/tables/geneSet_output.csv", row.names=FALSE)

options(width=130)
goDf[goDf$p.adjust < 0.05, c(1:5,7)]

################################################
#make heatmap of differentially expressed genes#
################################################

exprs_heatmap = vGene$E[rownames(sigGene),]

df <- as.data.frame(colData(rse_gene)[,c("PrimaryDx")])
rownames(df) <- colnames(exprs_heatmap)
colnames(df)<-"diagnosis"

pdf(file="DE_analysis/pdfs/de_heatmap.pdf")
pheatmap(exprs_heatmap, cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, annotation_col=df)
dev.off()
